Tail associations and their impact on extinction risk

### 2019-11-01 MainText.Rmd
working on discussion

<!--***DAN: May be scope for a discussion para linking what is known about
how ECEs are changing to what effect that may have on how tail assiciations
in environmental variables might be changing, and what influence that may have
on extinction risk. If little is known, then this might be a good topic for 
future work.-->

@@

<!--objectives, last para of intro-->
In this paper, we will explore the potential for asymmetric tail associations to 
influence extinction risk in populations and metapopulations. Specifically,
we will address the following sets of questions. Q1) Can the nature of tail associations
between vital rates in an age-structured population model influence
extinction risk for the model, and to what extent? Q2) Does the nature
of tail associations between measurements of an environmental variable in different
locations influence the extinction risk of a density-dependent metapopulation model
that depends on the environmental variable? How does the nature of the density
dependence mediate this influence? Will density dependence reverse or reproduce
the earlier result [@Ghosh_copula], using non-density-dependent models, that 
left-tail dependence in environmental variables accentuates and right-tail dependence 
mitigates extinction risks?
For Q1, we simulate a simple 2-stage model. We hypothesize that left-tail dependence
between juvenile survival and adult fecundity rates increases extinction risk
relative to right-tail dependence, even when the overall correlation is kept fixed. 
For Q2, we simulate spatial extensions of several simple and commonly
used population models. We intentionally use a simulation approach and simple models
instead of attempting mathematical analysis and/or using models that are more realistic, 
because our intention here is to provide an initial exploration only of 
the potential mechanisms, outlined above, which suggest that
asymmetries of tail association may make a difference for extinction risks. 
By so doing we hope to open these potentially practically important questions to 
wider examination other researchers. Some material in the Discussion outlines some possible next steps.
<!--***DAN: In Discussion need to talk about what analytic results are needed
to take the next step of determining when tail depencence may make enough of
a difference to think about. I think our results, together with Tulja's 
approximation, suggest that tail associations probably don't make much difference
for models without a spatial component. But our results illsutaret that it can
make a big difference when there is a spatial component, apparently because of 
the influence of tail associations on rescue effects. Could maybe even outline
analytic approaches for non-dd models? Not sure what would be done with dd
models. The point is you need to take the next step in the Discussion of 
talking about how our results, which admitedly are just for specific models,
could be turned into analytic results or general results. Interestingly, 
Tulja's approximation may also apply to the spatial case, at least when 
all locations can be reached from all other locations via dispersal. But the 
approximation seems poor in that case. Need to think about it more, and maybe
need more research.-->
Though tail associations are used
to assess risks in fields such as finance and hydrology 
[@alexander2008market; @li2013; @chen2011flood; @Goswami2018], to our
knowledge asymmetric tail associations have rarely been assessed for their
importance for extinction risks of populations and species (but see @Ghosh_copula). 
Our questions are an important step for understanding
the potential influence of ECEs on extinction risks if ECEs are related to
tail associations in environmental variables.

@@

<!--FIG : schematic diagram for partial spearman correlation, adapted from BIVAN-->
\begin{figure*}[!h]
\begin{center}
\includegraphics[width=6 cm]{./Results/PartialSpearmanFig.pdf}
\caption{The partial Spearman correlation, $\cor_{l_b,u_b}(u, v)$, within a band can be computed
for any band to describe how the strength of association between $u$ and $v$
varies from one part of the two distributions to another. Diagonal lines 
show two bands, the data in the right/upper band showing
stronger association than those in the left/lower band. This figure was reproduced
largely following Fig. 7 of Ghosh \emph{et al.} (2019).
\label{PartialSpearmanFig}}
\end{center}
\end{figure*}
<!--***DAN: Shyamolina, please change the u and v axis labels on the figure
to x and y, respectively, since we defined those as the ranks in the main text.-->

@@

To calculate extinction risk for a given metapopulation model, given parameters,
and a given choice of tail association,
we simulated the model for $25$ time steps, $10000$ times, using $n=5$ habitat patches.
The equilibrium value of the 
one-patch, deterministic version of the model, $P_e$ (Table \ref{tab_model_summary})
was used as the initial condition in all patches and simulations.
After each time step, if $P_i(t)$ was less than $P_e/10$ then it was set to $0$.
The whole metapopulation was considered to have gone extinct if all patches were $0$
at time step $25$.
<!--***DAN: Shyamolina, please check the above two sentences, is this what you did?
***Shya: I agreed with you upto line 482, the last line "The whole metapopulation was considered to have gone extinct if all patches were $0$ a time step $25$" I could not understand, did you write that conceptually? (then it's ok), 
or you meant I apply that in my code?
***DAN: I wrote it guessing that that was what you did. If that is not what you did,
when we should discuss, preferably as soon as possible, before Friday.
Similarly, the below line explains lines 149-150 of mydispersal.R, I am not fluent in using apply function on two dims at once e.g. c(1,3), you wrote that piece of code, I want to explain you 
on Friday gow I interpret that portion, then I can tell you accurately what I did.
***DAN: I added some comments above the function extrisk in mydispersal.R, after reading those
can you tell me by adding another comment what you did and if what I wrote here was correct?
-->
Extinction risk was the fraction of 10000 metapopulations that went extinct. 
Estimates of extinction risk were plotted together with 95\% confidence intervals
based on a binomial distribution.

<!--***DAN: Dan plans to delete this text as it has been superseded by the above. Keeping it
around until I am sure the above is right:
Our models (Table \ref{tab_model_summary}) 
computed extinction risk after $25$ years for $10000$ simulations at $n = 5$ patches
with spatially correlated noises having either moderate or extreme tail 
association in left and right-tail. 
We started with initial population $P_{t=0}^i = P_e^i, (i = 1,2,\ldots,n)$ for each patch, where 
$P_e^i$ was the equilibrium value (or similar to carrying capacity concept) 
for each density-dependence model (see 4$^{th}$ column of Table \ref{tab_model_summary}).
We assumed the population went extinct after each time step if it was below a 
certain threshold (set to 10\% of the initial population $P_e$ for individual model). 
-->

@@

<!--***DAN: stopped here, also need to do the caption of the last figure-->

@@

<!--summary results:-->
We showed the potential use of the "asymmetric" tail association in modeling the impact of 
extreme ecological events on population-extinction risk. Our results demonstrated left and right-tail 
association in vital rates (or in spatially correlated environmental noises) affected extinction risk differently
in years of extreme events ("catastrophes" and "bonanzas"). Though we could expect the left tail association 
to show higher extinction risk than the one for right tail association with age-structured 
and density-dependent metapopulation models, it was not always true. In the over-compensatory regime,
right tail dependent noise caused higher extinction risk than the left tail dependent noise. We showed 
that in the absence of dispersal the above-said switching took place near the critical parameter value where 
the deterministic model showed the transition from under-compensatory to the over-compensatory regime.

<!--Tuljapurkar approximation : linear version of age-structured model in stochastic environments-->
We found a left tail association in vital rates (induced by ECEs) caused higher extinction risk than 
the right tail dependent ones, but the difference was not remarkable 
unless the tail association was at its 
extreme (Fig. \ref{fig_extrisk_agestr}). This finding could be explained 
by the approximation @tuljapurkar1990lecture made for the 
stochastic log(growth rate) in an iid environment.
The approximation considered growth rate being 
dependent on the most dominant eigenvalue of the mean projection matrix 
and the temporal variance of the log(population growth rate) caused 
by environmental stochasticity - which was based
on the linear covariance measure between any 
two matrix elements, not on their tail association properties. Therefore,
following @tuljapurkar1990lecture, the population growth 
rate in an iid environment could be approximated
as same for our model (Eq. \ref{eq_age_str}) with 
left-tail or right-tail association.

@@

dependencies with copula could help in this case. Second, we did not consider multiple sites of 

@@

with spatially correlated extreme climatic events. Third, we considered the tail association 

@@

@@

interspecific-dependence. An earlier study [@carnicer2011widespread] showed extreme events

@@

a lot more potential to estimate the risk of cascading extinction in model food webs. <!--should I mention about 

@@

 <!--FIG : age structure model results : linear, nonlinear-->
\begin{figure*}[!h]
\textbf{ \hspace{5.4 cm} (A) \hspace{4 cm} (B)} \\
\vspace{-0.9 cm}
\begin{center}
\includegraphics[width=5 cm]{./Results/age_str_results/extrisk_age_str_C_SC_linear.pdf}
\includegraphics[width=5 cm]{./Results/age_str_results/extrisk_age_str_C_SC_nonlinear_omg_100.pdf}\\
\end{center}
\textbf{ \hspace{5.4 cm} (C) \hspace{4 cm} (D)} \\
\vspace{-0.9 cm}
\begin{center}
\includegraphics[width=5 cm]{./Results/age_str_results/extrisk_age_str_extremetail_linear.pdf}
\includegraphics[width=5 cm]{./Results/age_str_results/extrisk_age_str_extremetail_nonlinear_omg_100.pdf}
\caption{Extinction risk versus time for our stochastic, age-structured model; risks
were higher when associations between vital rates were stronger in the left compared to
the right tails of the rates, but only slightly so. Results are for the linear
version of the model ($\Omega = \infty$; A, C) and for the nonlinear model with density
dependence in the fecundity rate ($\Omega = 100$; B, D). Results show cases of moderate
(A, B) and extreme (C, D) asymmetry of tail association. The solid line is the estimated extinction risk and the area of shading, which is barely visible because it is so narrow, show 95\% confidence intervals for extinction risk.\label{fig_extrisk_agestr}}
\end{center}
\end{figure*}

<!--FIG :  summary : all nonlinear model results for local dispersion : extrisk vs. d-->
\begin{figure*}[!h]
\begin{center}
\includegraphics[width=14.5 cm]{./Results/allmodel_ext_risk_local_disp_extreme_and_moderate_noise.pdf}
\caption{Extinction risk versus dispersal rate for different scenarios of tail association
for our six metapopulation models (Methods, Table \ref{tab_model_summary}). Dispersal was
local, see Fig. \ref{SM-fig_extrisk_allmodel_globaldisp} for global dispersal. Values of $r$
used are in the panel headers.
Extinction risks are for a 25-year time horizon. \label{fig_extrisk_allmodel_localdisp}}
\end{center}
\end{figure*}

<!--FIG : all nonlinear model results, d=0, extrisk vs. r-->
\begin{figure*}[!h]
\begin{center}
\includegraphics[width=12 cm]{./Results/common_legend_extreme_and_moderate_noise.pdf}\\
\includegraphics[width=6 cm]{./Results/ricker_scl_1/ricker_extrisk_d0_extreme_and_moderate_noise.pdf}
\includegraphics[width=6 cm]{./Results/hassell_scl_1/hassell_extrisk_d0_extreme_and_moderate_noise.pdf}\\
\vspace{-0.6 cm}
\includegraphics[width=6 cm]{./Results/msmith_scl_0.8/msmith_extrisk_d0_extreme_and_moderate_noise.pdf}
\includegraphics[width=6 cm]{./Results/pennycuick_scl_1/pennycuick_extrisk_d0_extreme_and_moderate_noise.pdf}\\
\vspace{-0.6 cm}
\includegraphics[width=6 cm]{./Results/verhulst_scl_0.6/verhulst_extrisk_d0_extreme_and_moderate_noise.pdf}
\includegraphics[width=6 cm]{./Results/austinbrewer_scl_0.6/abrewer_extrisk_d0_extreme_and_moderate_noise.pdf}\\
\caption{Transition of higher to lower extinction risk (calculated after $25$ years) for 
the metapopulations affected by extreme left tail dependent noise across
$5$ multiple patches without dispersal ($d=0$), as intrinsic 
growth rate ($r$) goes from under to over-compensatory regime. Extinction risk for 
extreme right tail dependent noise follows the opposite trend.
Error bars represent the standard errors associated with $10000$
simulations for extinction risk computed at each $r$. Dotted line represents the value of $r$
for each model where their deterministic analog switches from under- to over-compensatory dynamics.
\label{fig_extrisk_d0}}
\end{center}
\end{figure*}

>>>>

@@

<!--objectives, last para of intro-->
In this paper, we will explore the potential for asymmetric tail associations to 
influence extinction risk in populations and metapopulations. Specifically,
we will address the following sets of questions. Q1) Can the nature of tail associations
between vital rates in an age-structured population model influence
extinction risk for the model, and to what extent? Q2) Does the nature
of tail associations between measurements of an environmental variable in different
locations influence the extinction risk of a density-dependent metapopulation model
that depends on the environmental variable? How does the nature of the density
dependence mediate this influence? Will density dependence reverse or reproduce
the earlier result [@Ghosh_copula], using non-density-dependent models, that 
left-tail dependence in environmental variables accentuates and right-tail dependence 
mitigates extinction risks?
For Q1, we simulate a simple 2-stage model. We hypothesize that left-tail dependence
between juvenile survival and adult fecundity rates increases extinction risk
relative to right-tail dependence, even when the overall correlation is kept fixed. 
For Q2, we simulate spatial extensions of several simple and commonly
used population models. We intentionally use a simulation approach and simple models
instead of attempting mathematical analysis and/or using models that are more realistic, 
because our intention here is to provide an initial exploration only of 
the potential mechanisms, outlined above, which suggest that
asymmetries of tail association may make a difference for extinction risks. 
By so doing we hope to open these potentially practically important questions to 
wider examination other researchers. Some material in the Discussion outlines some possible next steps.
Though tail associations are used
to assess risks in fields such as finance and hydrology 
[@alexander2008market; @li2013; @chen2011flood; @Goswami2018], to our
knowledge asymmetric tail associations have rarely been assessed for their
importance for extinction risks of populations and species (but see @Ghosh_copula). 
Our questions are an important step for understanding
the potential influence of ECEs on extinction risks if ECEs are related to
tail associations in environmental variables.

@@

<!--FIG : schematic diagram for partial spearman correlation, adapted from BIVAN-->
\begin{figure*}[!h]
\begin{center}
\includegraphics[width=6 cm]{./Results/PartialSpearmanFig.pdf}
\caption{The partial Spearman correlation, $\cor_{l_b,u_b}(u, v)$, within a band can be computed
for any band to describe how the strength of association between $u$ and $v$
varies from one part of the two distributions to another. Diagonal lines 
show two bands, the data in the right/upper band showing
stronger association than those in the left/lower band. This figure was reproduced
largely following Fig. 7 of Ghosh \emph{et al.} (2019).
\label{PartialSpearmanFig}}
\end{center}
\end{figure*}

@@

To calculate extinction risk for a given metapopulation model, given parameters,
and a given choice of tail association,
we simulated the model for $25$ time steps, $10000$ times, using $n=5$ habitat patches.
The equilibrium value of the 
one-patch, deterministic version of the model, $P_e$ (Table \ref{tab_model_summary})
was used as the initial condition in all patches and simulations.
After each time step, if $P_i(t)$ was less than $P_e/10$ then it was set to $0$.
The whole metapopulation was considered to have gone extinct if all patches were $0$
at time step $25$.
Extinction risk was the fraction of 10000 metapopulations that went extinct. 
Estimates of extinction risk were plotted together with 95\% confidence intervals
based on a binomial distribution.

@@

<!--***DAN: stopped here-->

@@

<!--summary results:-->
Our results combined with earlier results 
show that asymmetries in tail associations between environmental variables in
different locations can have a substantial influence on metapopulation extinction risk,
for density-dependent as well as for density-independent 
models. These results
extend the result of @Ghosh_copula, which was for a density-independent model. For a
density-independent model, @Ghosh_copula corroborated the intuition
that, when "harmful" environmental events 
(i.e., events which reduce the growth rates of local populations)
are correlated across space,
extinction risk of metapopulations will be accentuated compared 
to when "beneficial" events
(i.e., events which increase the growth of local populations)
are spatially correlated, even if overall correlation is the same in 
both scenarios. This was presumably because the first of these scenarios reduces the 
potential for rescue effects. For the density-dependent metapopulation 
models we considered, the same 
result held, typically, but not always, for undercompensatory model dynamics. However,
for overcompensatory dynamics, the oposite result held, again typically, but not always: 
when "harmful" events were correlated across space,
extinction risk of metapopulations was actually mitigated compared 
to when "beneficial" environmental events were 
spatially correlated by the same amount. This was probably because the correlated 
beneficial events caused populations to exceed carrying capacity, and 
to subsequently crash, across large areas. In contrast 
to the metapopulation case, our results 
suggest that asymmetry of tail associations between vital rates
in a stage-structured, non-spatial stochastic model may be of limited 
importance for extinction risk. However, our results provide suggestive, rather
than definitive evidence, here, because we considered 
only one model. In our stage-structured, non-spatial model, extinction 
risk was only weakly influenced by asymmetry of tail associations
between vital rates.

<!--***DAN: May be scope for a discussion para linking what is known about
how ECEs are changing to what effect that may have on how tail assiciations
in environmental variables might be changing, and what influence that may have
on extinction risk. If little is known, then this might be a good topic for 
future work.-->
Extensive evidence now exists that ECEs have become more frequent and more intense
over the last several decades, and substantial progress has recently been made so that
some event categories and specific events can now be attributed with confidence to
human-induced climate change [@ummenhofer2017extreme]. We here provide a fuller 
explanation of how such changes may modify the asymmetry of tail association of 
weather variables measured in different places, and thereby influence metapopulation
extinction risk through the mechansms that were explored in this paper.

<!--Tuljapurkar approximation : linear version of age-structured model in stochastic environments-->
<!--***DAN: Dan to elaborate and make more specific/precise-->
We found a left tail association in vital rates (induced by ECEs) caused higher extinction risk than 
the right tail dependent ones, but the difference was not remarkable 
unless the tail association was at its 
extreme (Fig. \ref{fig_extrisk_agestr}). This finding could be explained 
by the approximation @tuljapurkar1990lecture made for the 
stochastic log(growth rate) in an iid environment.
The approximation considered growth rate being 
dependent on the most dominant eigenvalue of the mean projection matrix 
and the temporal variance of the log(population growth rate) caused 
by environmental stochasticity - which was based
on the linear covariance measure between any 
two matrix elements, not on their tail association properties. Therefore,
following @tuljapurkar1990lecture, the population growth 
rate in an iid environment could be approximated
as same for our model (Eq. \ref{eq_age_str}) with 
left-tail or right-tail association.

<!--***DAN: Dan to maybe add another para on the fact that I thnk Tulja's approx should also
work for our spatial, non-dd model (check first, Dan), but we see a strong effect here,
so not clear why. And also suggests the possibilty that for some age-structured models,
tail associations will make a bigger difference then they did in our case. Maybe somehow
get to the suggestion that someone needs to revisit Tulja's approx and the analytic
approximation of extinction risk of Lande and Orzack looking at tail associations? Maybe
this and the previous para need to be combined, or broken up in a different way.-->

<!--***DAN: In Discussion need to talk about what analytic results are needed
to take the next step of determining when tail depencence may make enough of
a difference to think about. I think our results, together with Tulja's 
approximation, suggest that tail associations probably don't make much difference
for models without a spatial component. But our results illsutaret that it can
make a big difference when there is a spatial component, apparently because of 
the influence of tail associations on rescue effects. Could maybe even outline
analytic approaches for non-dd models? Not sure what would be done with dd
models. The point is you need to take the next step in the Discussion of 
talking about how our results, which admitedly are just for specific models,
could be turned into analytic results or general results. Interestingly, 
Tulja's approximation may also apply to the spatial case, at least when 
all locations can be reached from all other locations via dispersal. But the 
approximation seems poor in that case. Need to think about it more, and maybe
need more research.-->

@@

dependencies with copula could help in this case. 
<!--***DAN: the ideas in this para up to here (taking out the wrong statements) might 
follow on well from a para based on the notes I wrote above on Tulja's approx.-->
Second, we did not consider multiple sites of 

@@

with spatially correlated extreme climatic events. 
<!--***DAN: a reasonable point, above-->
Third, we considered the tail association 

@@

<!--***DAN: a good point, above-->

@@

interspecific-dependence. 
<!--***DAN: mentioning interactions strikes me as a step too far-->
An earlier study [@carnicer2011widespread] showed extreme events

@@

a lot more potential to estimate the risk of cascading extinction in model food webs. <!--should I 
mention about 

@@

<!--***DAN: might also mention how this relates with geog of sync of environmental vars, linking to
Jon's paper-->

 <!--FIG : age structure model results : linear, nonlinear-->
\begin{figure*}[!h]
\textbf{ \hspace{5.4 cm} (A) \hspace{4 cm} (B)} \\
\vspace{-0.9 cm}
\begin{center}
\includegraphics[width=5 cm]{./Results/age_str_results/extrisk_age_str_C_SC_linear.pdf}
\includegraphics[width=5 cm]{./Results/age_str_results/extrisk_age_str_C_SC_nonlinear_omg_100.pdf}\\
\end{center}
\textbf{ \hspace{5.4 cm} (C) \hspace{4 cm} (D)} \\
\vspace{-0.9 cm}
\begin{center}
\includegraphics[width=5 cm]{./Results/age_str_results/extrisk_age_str_extremetail_linear.pdf}
\includegraphics[width=5 cm]{./Results/age_str_results/extrisk_age_str_extremetail_nonlinear_omg_100.pdf}
\caption{Extinction risk versus time for our stochastic, age-structured model. Risks
were higher when associations between vital rates were stronger in the left compared to
the right tails of the rates, but only slightly so. Results are for the linear
version of the model ($\Omega = \infty$; A, C) and for the nonlinear model with density
dependence in the fecundity rate ($\Omega = 100$; B, D). Results show cases of moderate
(A, B) and extreme (C, D) asymmetry of tail association. The solid line is the estimated extinction risk, and the area of shading, which is barely visible because it is very narrow, shows 95\% confidence intervals for extinction risk.\label{fig_extrisk_agestr}}
\end{center}
\end{figure*}

<!--FIG :  summary : all nonlinear model results for local dispersion : extrisk vs. d-->
\begin{figure*}[!h]
\begin{center}
\includegraphics[width=14.5 cm]{./Results/allmodel_ext_risk_local_disp_extreme_and_moderate_noise.pdf}
\caption{Extinction risk versus dispersal rate for different scenarios of tail association
for our six metapopulation models (Methods, Table \ref{tab_model_summary}). Dispersal was
local, see Fig. \ref{SM-fig_extrisk_allmodel_globaldisp} for global dispersal. Values of $r$
used are in the panel headers.
Extinction risks are for a 25-year time horizon. Confidence intervals for extinction 
risk are not plotted, to provide a clearer plot. But 
because 10000 simulations were
done for each extinction risk estimate, confidence intervals
were very narrow, as in Fig. \ref{fig_extrisk_agestr}.\label{fig_extrisk_allmodel_localdisp}}
\end{center}
\end{figure*}

<!--FIG : all nonlinear model results, d=0, extrisk vs. r-->
\begin{figure*}[!h]
\begin{center}
\includegraphics[width=12 cm]{./Results/common_legend_extreme_and_moderate_noise.pdf}\\
\includegraphics[width=6 cm]{./Results/ricker_scl_1/ricker_extrisk_d0_extreme_and_moderate_noise.pdf}
\includegraphics[width=6 cm]{./Results/hassell_scl_1/hassell_extrisk_d0_extreme_and_moderate_noise.pdf}\\
\vspace{-0.6 cm}
\includegraphics[width=6 cm]{./Results/msmith_scl_0.8/msmith_extrisk_d0_extreme_and_moderate_noise.pdf}
\includegraphics[width=6 cm]{./Results/pennycuick_scl_1/pennycuick_extrisk_d0_extreme_and_moderate_noise.pdf}\\
\vspace{-0.6 cm}
\includegraphics[width=6 cm]{./Results/verhulst_scl_0.6/verhulst_extrisk_d0_extreme_and_moderate_noise.pdf}
\includegraphics[width=6 cm]{./Results/austinbrewer_scl_0.6/abrewer_extrisk_d0_extreme_and_moderate_noise.pdf}\\
\caption{Extinction risk versus $r$, with $d=0$, for different scenarios of tail association for 
our six metapopulation models (Methods, Table \ref{tab_model_summary}). 
Extinction risks are for a 25-year time horizon. Vertical lines show $r_c$,
the transition from under- to overcompensatory dynamics of the deterministic, one-patch
version of the model. Error bars show 95\% confidence intervals of extinction risk.
\label{fig_extrisk_d0}}
\end{center}
\end{figure*}
